Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-15 ieee_paper.tex
basic description of the swizzling

As explained in Section~XXX, we expect to achieve a performance of
about 210Gflop/s in single precision in the best case. We will focus
on the single precision case with 4 vectors and 4 matrices. But
similar techniques apply for other combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible without an efficient vectorization of
the kernel. Indeed, that means a Fused Multiply-Add instruction on  
fully loaded registers must be executed at most every 10 cycles.

Registers on the MIC are 512 bits wide, and can store 16 floats or 8
doubles. It is through these registers that the \#SIMD pragma is able
to achieve vectorization. (In comparison to OpenCL or CUDA, one can
think of these registers as a warp.)  We use the vector operations
presented in Figure~\ref{fig:MIC-vect-instr} in our code, which we'll
explain in the text.

@@

Here is a function that reads in four floats (a,b,c,d) and creates the 16-float vector dddd,cccc,bbbb,aaaa: 
\begin{verbatim}
__m512 read_aaaa(float* a)                                                              
{
    int int_mask_lo = (1 << 0) + (1 << 4) + (1 << 8) + (1 << 12);
    __mmask16 mask_lo = _mm512_int2mask(int_mask_lo);
    __m512 v1_old;
    v1_old = _mm512_setzero_ps();
    v1_old = _mm512_mask_loadunpacklo_ps(v1_old, mask_lo, a);
    v1_old = _mm512_mask_loadunpackhi_ps(v1_old, mask_lo, a);
    v1_old = _mm512_swizzle_ps(v1_old, _MM_SWIZ_REG_AAAA);
    return v1_old;
}
\end{verbatim}

Here is a function that takes four floats from float register \ttt{v1}, and 
places them in each of the four lanes. Because the permulation function only 
operates on 4-byte integers, it is necessary to convert (in place) each float
to an integer (the bit structure is not changed), permute the lanes and cast
the integers back to floats (with no modification of the bit structure)
\begin{verbatim}
__m512 permute(__m512 v1, _MM_PERM_ENUM perm)                                           
{
    __m512i vi = _mm512_castps_si512(v1);
    vi = _mm512_permute4f128_epi32(vi, perm);
    v1 = _mm512_castsi512_ps(vi);
    return v1;
}
\end{verbatim}

\newfig{figures/swizzling.pdf}{\wide}{Duplication of the first four
  bytes of each channel across the entire channel.}{fig:swizzling}

\newfig{figures/channel_permutation.pdf}{\wide}{Channel Permutations
  on the MIC}{fig:permutation}

\newfig{figures/tensor_product.pdf}{\wide}{Tensor scalar product using
  channels and swizzling.}{fig:tensor_product}

>>>>

As explained in Section~XXX, we expect to achieve a performance of
about 210Gflop/s in single precision in the best case. We will focus
on the single precision case with 4 vectors and 4 matrices. But
similar techniques apply for other combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible without an efficient vectorization of
the kernel. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. (We will show that an implementation relying on
compiler vectorization does not lead to desirable performance.) The
registers in the MIC architecture are 512 bits wide and can store 16
floats or integers. (In comparison to OpenCL or CUDA, one can think of
these registers as a warp.) The end goal is to be able to load and
format the non zero values $a,b,c,d$ and the vector entries $p,q,r,s$
in two vectorial register to be able to apply Fused Multiply-Add on
them, which is depicted in Figure~\ref{fig:tensor_product}. Bringing
the data in the vector register as depicted in the figure is the
difficult part that we now explain.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/tensor_product.pdf}
  \caption{Structure of the vector registers to perform an efficient
    tensor scalar product. See Figure~\ref{fig:mat_struct} for
    reference.}
  \label{fig:tensor_product}
\end{figure}

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of 4 lanes of 128 bits which are made of 4 elements
of 32 bits. A swizzling operation allows to reorder the elements
within each lane. While a permutation allow to change the orders of
the lanes. Figure~\ref{fig:swizzling_permutation} illustrates these
two possibilities. Both permutation and swizzling support common
reordering and broadcast.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/swizzling.pdf}
  \includegraphics[width=\linewidth]{figures/channel_permutation.pdf}

  \caption{Swizzling enables to reorder or broadcast elements within
    each 128 bit group. Permutation enables to reorder or broadcast
    while 128 bit groups.}
  \label{fig:swizzling_permutation}
\end{figure}

We use the vector operations presented in
Figure~\ref{fig:MIC-vect-instr} in our code, which we'll explain in
the text.

@@

Figure~\ref{code:read_aaaa} is a function that reads in four floats
(a,b,c,d) and creates the 16-float vector dddd,cccc,bbbb,aaaa:

\begin{figure*}
\begin{verbatim}
__m512 read_aaaa(float* a)                                                              
{
    int int_mask = 0x1111;
    __mmask16 mask = _mm512_int2mask(int_mask);
    __m512 v1;
    v1 = _mm512_setzero_ps();
    v1 = _mm512_mask_loadunpacklo_ps(v1, mask, a);
    v1 = _mm512_mask_loadunpackhi_ps(v1, mask, a);
    v1 = _mm512_swizzle_ps(v1, _MM_SWIZ_REG_AAAA);
    return v1_old;
}
\end{verbatim}
\caption{read\_aaaa code}
\label{code:read_aaaa}
\end{figure*}

Figure~\ref{code:permute} is a function that takes four floats from
float register \ttt{v1}, and places them in each of the four
lanes. Because the permulation function only operates on 4-byte
integers, it is necessary to convert (in place) each float to an
integer (the bit structure is not changed), permute the lanes and cast
the integers back to floats (with no modification of the bit
structure)

\begin{figure*}
\begin{verbatim}
__m512 permute(__m512 v1, _MM_PERM_ENUM perm)                                           
{
    __m512i vi = _mm512_castps_si512(v1);
    vi = _mm512_permute4f128_epi32(vi, perm);
    v1 = _mm512_castsi512_ps(vi);
    return v1;
}
\end{verbatim}
\caption{permute code}
\label{code:permute}
\end{figure*}
